---
title: "Topological Tree Distance"
output: html_document
date: "2023-12-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Topological Tree Distance using library apes

**Step 1:** Insert library

```{r}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

install.packages("treeio")
```

```{r}
install("treeio")
```

```{r}
library(ape)
library(treeio)
```

**Step 2:** Read trees from .nwk/.newick formats into document as objects of type "phylo"

```{r}
rocky <- read.tree(text = "((A:1, B:1):2, ((C:1, D:1):2, (E:1, F:1):2):4);")
plot(rocky)
```

```{r}
arabella <- read.tree(text = "((C:1, D:1):2, ((B:1, A:1):2, (E:1, F:1):2):4);")
plot(arabella)
```

```{r}
minnie <- read.tree(text = "((A:1, B:1):2, ((E:1, F:1):2, (C:1, D:1):2):4);")
plot(minnie)
```

```{r}
#read from file example, will return NULL is file is not formatted according to the specifications
filename <- "NewickExample.nwk"
barbarella <- read.tree(filename)
barbarella
```

```{r}
plot(barbarella)
```

```{r}
gorilla <- read.tree(text = "((A:2, B:2):4, ((C:2, D:2):4, (E:2, F:2):4):8);")
plot(gorilla)
```

**Step 3:** Calculate the topological distance between pairs of trees. The method, dist.topo, offers two different topological distance calculations.

**Method 1: Robinson and Foulds/Penny and Hendy**

In this method, the topological distance is calculated as 2 times the number of internal branches defining different bipartitions of the tips.

**Method 2: Branch length score (Kuhner and Felsenstein)**

In this method, the topological distance is calculated as the square root of the sum of the squared differences of the internal branch lengths defining similar bipartitions in both trees. Main difference is that this method takes branch lengths into account.

```{r}
#Method 1 on rocky and gorilla
dist.topo(rocky, gorilla, method = "PH85")
```

```{r}
#Method 2 on rocky and gorilla
dist.topo(rocky, gorilla, method = "score")
```

```{r}
#Method 1 on rocky and arabella
dist.topo(rocky, arabella, method = "PH85")
```

```{r}
#Method 2 on rocky and arabella
dist.topo(rocky, arabella, method = "score")
```

```{r}
#Method 1 on rocky and minnie
dist.topo(rocky, minnie, method = "PH85")
```

```{r}
#Method 2 on rocky and minnie
dist.topo(rocky, minnie, method = "score")
```

# **Import all trees**

The Bayesian MCMC tree is provided in nexus format.

The method read.nexus may produce an object of type "phylo" or "multiPhylo," so we are checking to make sure that a "phylo" object as being created (i.e., only one tree is being read).

```{r}
mcmc_tree_bad_tips <- read.nexus("BayesianMCMC.nexus")
class(mcmc_tree_bad_tips)
```

Confirm the tree has been correctly read by plotting it. Note how the tips contain labels with redundant information.

```{r}
plot(mcmc_tree_bad_tips)
```

Update tree tip labels to allow for analysis on the Bayesian MCMC tree.

```{r}
#step 1: generate new, desired tip names from old
mcmc_new_tips <- vector(mode='character',length=21)
for(i in 1:21)
{
  mcmc_new_tips[i] <- paste(substring(mcmc_tree_bad_tips$tip.label[i], 1, 9), substring(mcmc_tree_bad_tips$tip.label[i], 57, 66), sep = "")
}
mcmc_new_tips
```

```{r}
d <- data.frame(mcmc_tree_bad_tips$tip.label, mcmc_new_tips)
mcmc_tree <- rename_taxa(mcmc_tree_bad_tips, d)
plot(mcmc_tree)
```

The UPGMA, Neighbor-Joining, Parsimony, and MLE trees are provided in newick format.

```{r}
upgma_tree <- read.tree("UPGMATree.nwk")
neighbor_joining_tree <- read.tree("NJTree.nwk")
parsimony_tree <- read.tree("parsimony.nwk")
mle_tree <- read.tree("MLE.nwk")
```

Once again confirm that each tree is correctly read by plotting it to check for tip labels.

```{r}
plot(upgma_tree)
```

```{r}
plot(neighbor_joining_tree)
```

```{r}
plot(parsimony_tree)
```

```{r}
plot(mle_tree)
```

This analysis requires the use of unrooted trees, so we will make sure that all trees are unrooted before running the analysis.

```{r}
is.rooted(mcmc_tree)
mcmc_tree <- unroot(mcmc_tree)
is.rooted(mcmc_tree)
```

```{r}
is.rooted(upgma_tree)
upgma_tree <- unroot(upgma_tree)
is.rooted(upgma_tree)
```

```{r}
is.rooted(neighbor_joining_tree)
neighbor_joining_tree <- unroot(neighbor_joining_tree)
is.rooted(neighbor_joining_tree)
```

```{r}
is.rooted(parsimony_tree)
parsimony_tree <- unroot(parsimony_tree)
is.rooted(parsimony_tree)
```

```{r}
is.rooted(mle_tree)
```

# Topological Distances Calculations

Define function to call topological distance calculations on each pair of trees.

```{r}
trees <- c(mcmc_tree, upgma_tree, neighbor_joining_tree, parsimony_tree, mle_tree)

calculate_topo_dists <- function(tree, calc_method){
  output <- c(1,1,1,1,1)
  index = 1
  for (cmp_tree in trees)
  {
    output[index] = dist.topo(tree, cmp_tree, method = calc_method)
    index = index + 1
  }
  return(output)
}
```

Calculate all pairs of topological distances.

```{r}
topo_dists_table <- function(calc_method) {
  mcmc_topo_dists <- calculate_topo_dists(mcmc_tree, calc_method)
  upgma_topo_dists <- calculate_topo_dists(upgma_tree, calc_method)
  neighbor_joining_topo_dists <- calculate_topo_dists(neighbor_joining_tree, calc_method)
  parsimony_dists <- calculate_topo_dists(parsimony_tree, calc_method)
  mle_topo_dists <- calculate_topo_dists(mle_tree, calc_method)
  
  #topo_dists_vector <- c(mcmc_topo_dists, upgma_topo_dists, neighbor_joining_topo_dists, parsimony_dists, mle_topo_dists)
  #topo_dists = matrix(topo_dists_vector, nrow = 5, ncol = 5)
  topo_dists = data.frame(Methods = c("Bayesian_MCMC", "UPGMA", "NJ", "Parsimony", "MLE"),
    Bayesian_MCMC = mcmc_topo_dists,
                          UPGMA = upgma_topo_dists,
                          NJ = neighbor_joining_topo_dists,
                          Parsimony = parsimony_dists,
                          MLE = mle_topo_dists, row.names = 1)
  return(topo_dists)
}
```

Place all topological distances in a 2 x 2 vector.

```{r}
topo_dists_RF <- topo_dists_table("PH85")
topo_dists_branch_length_score <- topo_dists_table("score")
```

Pretty print all topological distances

```{r}
topo_dists_RF
```

```{r}
topo_dists_branch_length_score
```

Finally, write these outputs to csv files.

```{r}
write.csv(topo_dists_RF, "RobinsonFouldsTopologicalDistances.csv", row.names=TRUE)
```

```{r}
write.csv(topo_dists_branch_length_score, "BranchLengthScoreTopologicalDistances.csv", row.names=TRUE)
```
